A Dependently Typed Multi-Stage Calculus \ifnotanonymous\thanks{Supported by organization x.}\fi

### 2019-06-15 text/formal-defnition.tex
Revised almost all comments.

\AI{Is $O$ used as a metavariable?}
\AI{Domain and FV are not defined (yet).  I think we should introduce the notion of well-formed type environments by prose below and assume every type environment is well formed.}

@@

We adopt a \LLF-like system, where constants and type-level constants are given in the signature $\Sigma$, which
are a sequence of declarations of the form $c:\tau$ and $X:K$.
For example, when we use integers in \LMD, $\Sigma = \B :: *, true:\B, false:\B$.
Please note that constants or type-level constants has no stage.
This means that we can use them at any stage.
% Type environments
Type environments are sequences of triples of a variable, its type, and its stage.
We define well-formed signature and well-formed type environments later
in order to prove theorems.

@@

Next, we define full reduction for \LMD.
Before giving the definition of reduction, we define two kinds of substitutions.
Substitution $M[x\mapsto N]$ is the ordinary capture-avoiding substitution of
$N$ for $x$ in $M$, and we omit its definition here.
Substitution $M[\alpha \mapsto \varepsilon]$ of the empty stage
for stage variable $\alpha$ in $M$ is defined below.
\AI{What about $\tau[x \mapsto N], K[x \mapsto N], \tau [\alpha \mapsto \varepsilon]$ and $K[\alpha \mapsto \varepsilon]$?}

\AI{Since we always substitute $\varepsilon$ for stage variables, we may want to introduce a special notation, something like $\textit{erase}_\alpha(M)$.}
\AI{We will use $[\alpha \mapsto \varepsilon]$ and $[\alpha \mapsto \beta]$.  The latter is renaming, so we can omit the definition.}

\begin{align*}
	% \newcommand{\SB}{}
	(\lambda x:\tau.M)[\alpha \mapsto \varepsilon] & = \lambda x:(\tau[\alpha \mapsto \varepsilon]).(M[\alpha \mapsto \varepsilon])                                  \\
	(M\ N)[\alpha \mapsto \varepsilon]             & = (M[\alpha \mapsto \varepsilon])\ (N[\alpha \mapsto \varepsilon])                                              \\
	(M\ \varepsilon)[\alpha \mapsto \varepsilon]   & = (M[\alpha \mapsto \varepsilon])\ \varepsilon                                                                  \\
	(\TB_\alpha M)[\alpha \mapsto \varepsilon]     & = M[\alpha \mapsto \varepsilon]                                                                                 \\
	(\TB_\beta M)[\alpha \mapsto \varepsilon]      & = \TB_\beta (M[\alpha \mapsto \varepsilon])                                    & (\text{if } \alpha \neq \beta) \\
	(\TBL_\alpha M)[\alpha \mapsto \varepsilon]    & =  M[\alpha \mapsto \varepsilon]                                               &                                \\
	(\TBL_\beta M)[\alpha \mapsto \varepsilon]     & = \TBL_\beta (M[\alpha \mapsto \varepsilon])                                   & (\text{if } \alpha \neq \beta) \\
	(\Lambda\beta.M)[\alpha \mapsto \varepsilon]   & = \Lambda\beta.(M[\alpha \mapsto \varepsilon])                                 & (\text{if } \alpha \neq \beta) \\
	% (\Lambda\beta.M)[\alpha \mapsto \varepsilon]   & = \Lambda\beta.M                                                     & (\text{if } \alpha = \beta)    \\
	(\%_\alpha M)[\alpha \mapsto \varepsilon]      & = M[\alpha \mapsto \varepsilon]                                                &                                \\
	(\%_\beta M)[\alpha \mapsto \varepsilon]       & = \%_{\beta[\alpha \mapsto \varepsilon]}(M[\alpha \mapsto \varepsilon])        & (\text{if } \alpha \neq \beta) \\
\end{align*}

\begin{definition}[Reduction]
	\AI{It's not clear what are defined here.  You are talking only about rules.
		Something like ``The relations $M \longrightarrow_\beta N$, ... are the least compatible relations closed under the rules below.''}
	There are three reduction rules ($\longrightarrow_\beta, \longrightarrow_\blacklozenge, \longrightarrow_\Lambda$) in \LMD.
	Congruence rules which are omitted from the definition.
	\begin{align*}
		 & (\lambda x:\tau.M) N \longrightarrow_\beta M[x \mapsto N]                             \\
		% & (\Pi x:\tau.\sigma) M \longrightarrow_\gamma \sigma[x \mapsto M] \\
		 & \TBL_\alpha \TB_\alpha M \longrightarrow_\blacklozenge M                              \\
		 & (\Lambda \alpha.M)\ \varepsilon \longrightarrow_\Lambda M[\alpha \mapsto \varepsilon]
	\end{align*}
	We write $ M \longrightarrow M'$ iff $ M \longrightarrow_\beta M'$, $ M \longrightarrow_\blacklozenge M'$, or $ M \longrightarrow_\Lambda M'$.
\end{definition}

@@

In this section, we define the type system of \LMD.
% \LMD is little complicated because it contains dependent types.
It consists of six judgment forms for typing, kinding, well-formed kinding, term equivalence, type equivalence, and kind equivalence.
We list the judgments forms in Figure~\ref{fig:LMD-six-judgments}.
\AI{equality vs. equivalence?}
\AI{A few judgment forms do not match rules.}

@@

\AI{Insert a tilde (which is a space without a line break) between ``Figure'' and ref.}
\begin{definition}[Typing, Kinding, Well-formed Kinding, Term Equivalence, Type Equivalence, Kind Equivalence]
	The typing relation $ \G \V M : \tau @ A $,
	the kinding relation $\G \V \tau :: K @ A$,
	the well-formed kinding relation $\G \V K \iskind @ A$,
	the term equivalence relation $\G \V M \E N @ A$,
	the type equivalence relation $\G \V \tau \E \sigma @ A$,
	the kind equivalence relation $\G \V K \E J @ A$,
	are the least relation closed under the rules in Figure~\ref{fig:typing-rules}.
\end{definition}
This definition is little complicated to define six relations at the same time.
We canot define them one by one because they are dependent on each other.

@@

 We start from defining typing of
\LMD.  Typing rules of \LMD are shown in Figure
\ref{fig:typing-rules}.  The rule \TConst{} means any constants in the
signature can appear at any stage.  \AI{This should be discussed
	earlier.}  For example, if we have a signature $\Sigma$ which is
$\textrm{bool} :: *, \textrm{true}: \textrm{bool}, \textrm{false}:
	\textrm{bool}$, the derivation tree in
Figure~\ref{fig:tconst-derivation-tree} is admissible.  \AI{Don't we
	need signature well formedness?  If $c:\tau \in \Sigma$, then $\tau$
	should be a well-formed type under ... what?}  The rules \TVar,
\TAbs, and \TApp{} are almost the same as those in the simply typed
lambda calculus.  \AI{... or \(\lambda\)LF?  We are working in a dependently type system...}  Additional conditions are that subterms must be
typed at the same stage (\TAbs{} and \TApp); the type
annotation/declaration on a variable has to be a proper type of kind
$*$ (\TAbs) at the stage where it is declared (\TVar{} and \TAbs).

@@

The term equivalence judgment of the form $\G \V M \E N : \rho @ A$,
which means that terms $M$ and $N$ are equivalent as terms of type
$\rho$ at stage $A$ under $\G$, is defined by the rules in
Figure~\ref{fig:term-equivalence-rules}.  Most rules are
straightforward.  The rules \QAbs, \QApp, \QTB, \QTBL, \QGen, \QIns,
\QCsp, \QRefl, \QSym, and \QTrans{} make the relation congruence; the
rules \QBeta, \QTBLTB, and \QLambda{} correspond to the reduction rules.
\AI{Don't we give names to those reduction rules?}

@@

This kind of type equality is very useful when we combine $\text{mulmat}\ 3\ 5$ with another code.
We cannot remove $\%_\alpha$ in the type without \QPercent and it makes very difficult to combine
because types of codes don't contain $\%_\alpha$ generally.

>>>>

\AI{Is $O$ used as a metavariable?}
% \AI{Domain and FV are not defined (yet).  I think we should introduce the notion of well-formed type environments by prose below and assume every type environment is well formed.}

@@

We adopt a \LLF-like system, where constants and type-level constants are given in the signature $\Sigma$, which
are a sequence of declarations of the form $c:\tau$ and $X:K$.
For example, when we use integers in \LMD, $\Sigma = \B :: *, \textrm{true}:\B, \textrm{false}:\B$.
Please note that constants or type-level constants has no stage.
This means that we can use them at any stage.
% Type environments
Type environments are sequences of triples of a variable, its type, and its stage.
We define well-formed signature and well-formed type environments later.

@@

Next, we define full reduction for \LMD.
Before giving the definition of reduction, we define six kinds of substitutions.
Substitution $M[x\mapsto N], \tau[x \mapsto N]$ and $K[x \mapsto N]$ are 
the ordinary capture-avoiding substitutions of
term $N$ for $x$ in term $M$, type $\tau$, and kind $K$, respectively
and we omit their definition here.
Substitution $M[\alpha \mapsto A], \tau [\alpha \mapsto A], K[\alpha \mapsto A]$ and $B[\alpha\mapsto A]$ are 
substitutions of stage $A$ for stage variable $\alpha$ in 
term $N$ for $x$ in term $M$, type $\tau$, kind $K$, and stage $B$ respectively.
We show replesentive cases below.
% \AI{What about $\tau[x \mapsto N], K[x \mapsto N], \tau [\alpha \mapsto \varepsilon]$ and $K[\alpha \mapsto \varepsilon]$?}

% \AI{Since we always substitute $\varepsilon$ for stage variables, we may want to introduce a special notation, something like $\textit{erase}_\alpha(M)$.}
% \AI{We will use $[\alpha \mapsto \varepsilon]$ and $[\alpha \mapsto \beta]$.  The latter is renaming, so we can omit the definition.}

\begin{align*}
	(\lambda x:\tau.M)[\alpha \mapsto A] & = \lambda x:(\tau[\alpha \mapsto A]).(M[\alpha \mapsto A])                                  \\
	(M\ B)[\alpha \mapsto A]             & = (M[\alpha \mapsto A])\ B[\alpha\mapsto A]                                                 \\
	(\TB_\beta M)[\alpha \mapsto A]      & = \TB_{\beta[\alpha \mapsto A]}M[\alpha \mapsto A]                                          \\
	(\TBL_\beta M)[\alpha \mapsto A]     & = \TBL_{\beta[\alpha \mapsto A]}M[\alpha \mapsto A]                                         \\
	(\%_\beta M)[\alpha \mapsto A]       & = \%_{\beta[\alpha \mapsto A]}M[\alpha \mapsto A]                                           \\
	(\beta B)[\alpha \mapsto A]          & = \beta (B[\alpha\mapsto A])                               & (\text{if } \alpha \neq \beta) \\
	(\beta B)[\alpha \mapsto A]          & = A (B[\alpha\mapsto A])                                   & (\text{if } \alpha = \beta)
\end{align*}

\begin{definition}[Reduction]
	The relations $M \longrightarrow_\beta N$, $M \longrightarrow_\blacklozenge N$, and $M \longrightarrow_\Lambda N$
	are the least compatible relations closed under the rules below.
	Congruence rules which are omitted from the definition.
	\begin{align*}
		 & (\lambda x:\tau.M) N \longrightarrow_\beta M[x \mapsto N]                             \\
		 & \TBL_\alpha \TB_\alpha M \longrightarrow_\blacklozenge M                              \\
		 & (\Lambda \alpha.M)\ A \longrightarrow_\Lambda M[\alpha \mapsto A]
	\end{align*}
	We write $ M \longrightarrow M'$ iff $ M \longrightarrow_\beta M'$, $ M \longrightarrow_\blacklozenge M'$, or $ M \longrightarrow_\Lambda M'$ and
	we call $\longrightarrow_\beta$, $\longrightarrow_\blacklozenge$, and $\longrightarrow_\Lambda$ $\beta$-reduction, $\blacklozenge$-reduction, and $\Lambda$-reduction, respectively.
\end{definition}

@@

In this section, we define the type system of \LMD.
% \LMD is little complicated because it contains dependent types.
It consists of six judgment forms for typing, kinding, well-formed kinding, term equivalence, type equivalence, and kind equivalence.
We list the judgments forms in Figure~\ref{fig:LMD-six-judgments}.
% \AI{equality vs. equivalence?} \red{Select equivalence}
% \AI{A few judgment forms do not match rules.}

@@

% \AI{Insert a tilde (which is a space without a line break) between ``Figure'' and ref.}
\begin{definition}[Typing, Kinding, Well-formed Kinding, Term Equivalence, Type Equivalence, Kind Equivalence]
	The typing relation $ \G \V M : \tau @ A $,
	the kinding relation $\G \V \tau :: K @ A$,
	the well-formed kinding relation $\G \V K \iskind @ A$,
	the term equivalence relation $\G \V M \E N @ A$,
	the type equivalence relation $\G \V \tau \E \sigma @ A$,
	the kind equivalence relation $\G \V K \E J @ A$,
	are the least relation closed under the rules in Figure~\ref{fig:typing-rules}.
\end{definition}
This definition is little complicated to define six relations at the same time.
We canot define them one by one because they are dependent on each other.

@@

We start from defining typing of
\LMD.  Typing rules of \LMD are shown in Figure~\ref{fig:typing-rules}.
The rule \TConst{} means any constants in the
signature can appear at any stage.  \
AI{This should be discussed earlier.}  
For example, if we have a signature $\Sigma$ which is
$\textrm{bool} :: *, \textrm{true}: \textrm{bool}, \textrm{false}:
	\textrm{bool}$, the derivation tree in
Figure~\ref{fig:tconst-derivation-tree} is admissible.  
The rules \TVar,\TAbs, and \TApp{} are almost the same as those in the simply typed
lambda calculus or \LLF.  Additional conditions are that subterms must be
typed at the same stage (\TAbs{} and \TApp); the type
annotation/declaration on a variable has to be a proper type of kind
$*$ (\TAbs) at the stage where it is declared (\TVar{} and \TAbs).
% \AI{Don't we need signature well formedness?  If $c:\tau \in \Sigma$, then $\tau$ should be a well-formed type under ... what?}
% \AI{... or \(\lambda\)LF?  We are working in a dependently type system...}

@@

The term equivalence judgment of the form $\G \V M \E N : \rho @ A$,
which means that terms $M$ and $N$ are equivalent as terms of type
$\rho$ at stage $A$ under $\G$, is defined by the rules in
Figure~\ref{fig:term-equivalence-rules}.  Most rules are
straightforward.  The rules \QAbs, \QApp, \QTB, \QTBL, \QGen, \QIns,
\QCsp, \QRefl, \QSym, and \QTrans{} make the relation congruence; the
rules \QBeta, \QTBLTB, and \QLambda{} correspond to 
$\beta$-reduction, $\blacklozenge$-reduction, and $\Lambda$-reduction, respectively.

@@

This kind of type equivalence is very useful when we combine $\text{mulmat}\ 3\ 5$ with another code.
We cannot remove $\%_\alpha$ in the type without \QPercent and it makes very difficult to combine
because types of codes don't contain $\%_\alpha$ generally.
